\name: Deterministic and Coherent Implicits

### 2018-04-19 jfp/src/Translation.tex
adding hyperlinks

%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
\paragraph{Resolution}
The more interesting part of the translation happens when resolving queries.
Queries are translated by rule $\TyQuery$ using the auxiliary resolution
judgment $\ivturns$:
\begin{equation*}
\tenv \ivturns \rulet~\gbox{\leadsto E}
\end{equation*}
which is shown, in deterministic form, in Figure~\ref{fig:resolution2}.  The
translation builds a System F term as evidence for the resolution.

The mechanism that builds evidence dualizes the process of peeling off
abstractions and universal quantifiers: Rule \mylabel{R-IAbs}~wraps a lambda
binder with a fresh variable $x$ around a System F expression $E$, which is
generated from the resolution for the head of the rule ($\rulet_2$). Similarly,
rule \mylabel{R-TAbs}~wraps a type lambda binder around the System F expression
resulting from the resolution of $\rulet$.

For simple types $\type$ rule \mylabel{R-Simp} delegates the work of
building evidence, when a matching rule $\rulet$ type is found in the
environment, to rule \mylabel{L-RuleMatch}. The evidence consists of two parts:
$E$ is the evidence of matching $\type$ against $\rulet$. This match contains
placeholders $\bar{x}$ for the contexts whose resolution is postponed by rule
\mylabel{M-IAbs}. It falls to rule \mylabel{L-RuleMatch} to perform these
postponed resolutions, obtain their evidence $\bar{E}$ and fill in the
placeholders.

@@

Moreover, we can express three key properties of Figure~\ref{fig:resolution2}'s
definition of resolution in terms of the generated evidence.
Firstly, the deterministic version of resolution is a sound variation on the original ambiguous resolution.
\begin{lemma}[Soundness]
Figure~\ref{fig:resolution2}'s definition of resolution is sound (but
incomplete) with respect to Figure~\ref{fig:resolution1}'s definition.
\[\forall \tenv, \rulet, E: \quad\quad \tenv \ivturns \rulet \leadsto E \quad\Rightarrow\quad \tenv \vturns \rulet \leadsto E \]
\end{lemma}
Secondly, the deterministic resolution guarantees a strong form of coherence:
\begin{lemma}[Determinacy]
The generated evidence of resolution is uniquely determined.
\[\forall \tenv, \rulet, E_1, E_2: \quad\quad \tenv \ivturns \rulet \leadsto E_1 ~\wedge~ \tenv \ivturns \rulet \leadsto E_2 \quad\Rightarrow\quad E_1 = E_2 \]
\end{lemma}
Thirdly, on top of the immediate coherence of deterministic resolution, 
an additional coherence property holds.  
\begin{lemma}[Stability]
Resolution is stable under substitution.
\[\forall \tenv,\alpha,\tenv',\sigma,\rulet, E: \enskip\
\tenv,\alpha,\tenv' \ivturns \rulet \leadsto E \enskip\wedge\enskip \tenv \vdash \sigma
\enskip\Rightarrow\enskip 
\tenv,\tenv'[\sigma/\alpha] \ivturns \rulet[\sigma/\alpha] \leadsto E[|\sigma|/\alpha] \]
\end{lemma}

@@

% The main difference, and complication, is due to the fact that the evidence for
% type instantiation and recursive resolution is needed before these operations
% actually take place, as the algorithm has to postpone them. For this reason, the
% algorithm first produces placeholders that are later substituted for the actual
% evidence.
% 
% The central relation is $\rulet; \bar{\rulet}; \bar{\alpha} \gbox{; \bar{\omega};
% E}\turns_{\mathit{match}} \tau \hookrightarrow \bar{\rulet}' \gbox{;
% \bar{\omega}'; E'}$. It captures the matching instantiation of context type
% $\rulet$ against simple type $\tau$.  The input evidence for $\rulet$ is $E$, and
% the output evidence for the instantiation is $E'$. The accumulating parameters
% $\bar{\alpha}$ and $\bar{\rulet}$ denote that the instantiation of type variables
% $\bar{\alpha}$ and the recursive resolution of $\bar{\rulet}$ have been
% postponed. We use the $\bar{\alpha}$ themselves as convenient placeholders for
% the instantiating types, and we use the synthetic $\bar{\omega}$ as placeholders for
% the evidence of the $\bar{\rulet}$. The rules \mylabel{MTC-Abs} and \mylabel{MTC-Abs} 
% introduce these two kinds of placeholders in the evidence. The former kind, $\bar{\alpha}$,
% are substituted in rule \mylabel{MTC-Simp} where the actual type instantiatons $\bar{\theta}$
% are computed. The latter kind, $\bar{\omega}$, are substituted later in rule \mylabel{Alg-Simp} where
% the recursive resolutions take place.

>>>>

%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
\paragraph{Resolution}
The more interesting part of the translation happens when resolving queries.
Queries are translated by rule $\rref{Ty-Query}$ using the auxiliary resolution
judgment $\ivturns$:
\begin{equation*}
\tenv \ivturns \rulet~\gbox{\leadsto E}
\end{equation*}
which is shown, in deterministic form, in Figure~\ref{fig:resolution2}.  The
translation builds a System F term as evidence for the resolution.

The mechanism that builds evidence dualizes the process of peeling off
abstractions and universal quantifiers: Rule \rref{R-IAbs}~wraps a lambda
binder with a fresh variable $x$ around a System F expression $E$, which is
generated from the resolution for the head of the rule ($\rulet_2$). Similarly,
rule \rref{R-TAbs}~wraps a type lambda binder around the System F expression
resulting from the resolution of $\rulet$.

For simple types $\type$ rule \rref{R-Simp} delegates the work of
building evidence, when a matching rule $\rulet$ type is found in the
environment, to rule \rref{L-RuleMatch}. The evidence consists of two parts:
$E$ is the evidence of matching $\type$ against $\rulet$. This match contains
placeholders $\bar{x}$ for the contexts whose resolution is postponed by rule
\rref{M-IAbs}. It falls to rule \rref{L-RuleMatch} to perform these
postponed resolutions, obtain their evidence $\bar{E}$ and fill in the
placeholders.

@@

Moreover, we can express three key properties of Figure~\ref{fig:resolution2}'s
definition of resolution in terms of the generated evidence.
Firstly, the deterministic version of resolution is a sound variation on the original ambiguous resolution.
\begin{lemma}[Soundness]
Figure~\ref{fig:resolution2}'s definition of resolution is sound (but
incomplete) with respect to Figure~\ref{fig:resolution1}'s definition.
\[\forall \tenv, \rulet, E: \quad\quad \tenv \ivturns \rulet \leadsto E \quad\Rightarrow\quad \tenv \vturns \rulet \leadsto E \]
\end{lemma}
Secondly, the deterministic resolution guarantees a strong form of coherence:
\begin{lemma}[Determinacy]
The generated evidence of resolution is uniquely determined.
\[\forall \tenv, \rulet, E_1, E_2: \quad\quad \tenv \ivturns \rulet \leadsto E_1 ~\wedge~ \tenv \ivturns \rulet \leadsto E_2 \quad\Rightarrow\quad E_1 = E_2 \]
\end{lemma}
Thirdly, on top of the immediate coherence of deterministic resolution, 
an additional coherence property holds.  
\begin{lemma}[Stability]
Resolution is stable under type substitution.
\[\forall \tenv,\alpha,\tenv',\sigma,\rulet, E: \enskip\
\tenv,\alpha,\tenv' \ivturns \rulet \leadsto E \enskip\wedge\enskip \tenv \vdash \sigma
\enskip\Rightarrow\enskip 
\tenv,\tenv'[\sigma/\alpha] \ivturns \rulet[\sigma/\alpha] \leadsto E[|\sigma|/\alpha] \]
\end{lemma}

@@

% The main difference, and complication, is due to the fact that the evidence for
% type instantiation and recursive resolution is needed before these operations
% actually take place, as the algorithm has to postpone them. For this reason, the
% algorithm first produces placeholders that are later substituted for the actual
% evidence.
% 
% The central relation is $\rulet; \bar{\rulet}; \bar{\alpha} \gbox{; \bar{\omega};
% E}\turns_{\mathit{match}} \tau \hookrightarrow \bar{\rulet}' \gbox{;
% \bar{\omega}'; E'}$. It captures the matching instantiation of context type
% $\rulet$ against simple type $\tau$.  The input evidence for $\rulet$ is $E$, and
% the output evidence for the instantiation is $E'$. The accumulating parameters
% $\bar{\alpha}$ and $\bar{\rulet}$ denote that the instantiation of type variables
% $\bar{\alpha}$ and the recursive resolution of $\bar{\rulet}$ have been
% postponed. We use the $\bar{\alpha}$ themselves as convenient placeholders for
% the instantiating types, and we use the synthetic $\bar{\omega}$ as placeholders for
% the evidence of the $\bar{\rulet}$. The rules \rref{MTC-Abs} and \rref{MTC-Abs} 
% introduce these two kinds of placeholders in the evidence. The former kind, $\bar{\alpha}$,
% are substituted in rule \rref{MTC-Simp} where the actual type instantiatons $\bar{\theta}$
% are computed. The latter kind, $\bar{\omega}$, are substituted later in rule \rref{Alg-Simp} where
% the recursive resolutions take place.
